Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-11-24 concurrency/CPP 2024/paper.tex
resolved some TODO

%But first, we need to understand how to specify the correctness of \lstinline{traverse} independently of the data structure operations (insert, lookup, etc.) it will be used to implement. The key is the concept of \emph{logical atomicity}~\cite{tada}, a highly compositional approach to specifying concurrent data structure operations. \wm{Should this go in background instead?}

Logical atomicity is a separation logic technique for concisely specifying the behavior of a concurrent operation. A logically atomic triple has the form $\fforall a.\,\left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle\ \texttt{c}\ \left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, this says that \lstinline{c} is an operation on an abstract object (i.e., data structure) $a$, and its effect is to \emph{atomically} transform $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread. More precisely, the triple asserts that if $\texttt{P}_l$ holds true before a call to \lstinline{c} and $\texttt{P}_p$ is true for some value of $a$ in a shared state, then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends).

@@

\begin{figure*}[!ht]
	$\color{specblue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{specblue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

\begin{figure*}[h]
	\begin{subfigure}{\textwidth}
		\centering
		\begin{mathpar}
			{\color{specblue}
				\left\{ 
				\begin{array}{c}
					\ \texttt{pn->n} \mapsto \texttt{NULL}  
				\end{array}
				\right\}
			} \vspace{-0.85em} 
			\\ 
			\texttt{insertOp(pn, k, v)} 
			\vspace{-0.85em} 
			\\
			{\color{specblue}
				\left\{\exists \  l\ r\ \mathit{lk}_1\ \mathit{lk}_2.
				\begin{array}{l}
					\ \texttt{pn->n->t} \mapsto (\texttt{k}, \texttt{v}, l, r) \ \ast \ \\ 
					\ l \mapsto (\texttt{NULL}, \mathit{lk}_1)  \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2)
				\end{array}
				\right\}
			}
		\end{mathpar}
		\caption{Specification of \texttt{insertOp} for the lock-coupling template}
		\label{fig:insertOp_lock}
	\end{subfigure}
	%insertOp for give-up of give-up
	\begin{subfigure}{\textwidth}
		\begin{mathpar}
			{\color{specblue}
				\left\{ 
				\begin{array}{c}
					\texttt{pn->n}\mapsto \texttt{NULL}  \ast  \texttt{pn->n->min}\mapsto n_1  \ast  \texttt{pn->n->max}\mapsto n_2
				\end{array}
				\right\}
			} \vspace{-0.85em} 
			\\
			\texttt{insertOp(pn, k, v)}
			\vspace{-0.85em} 
			\\
			{\color{specblue}
				\left\{\exists \ l\ r\ \mathit{lk}_1\ \mathit{lk}_2. 
				\begin{array}{c}
					\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r)  \ast  \texttt{pn->n->min}\mapsto n_1 \ast  \texttt{pn->n->max}\mapsto n_2  \ \ast \ 
					\\ l \mapsto (\texttt{NULL}, \mathit{lk}_1, (n_1, \texttt{k})) \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2, (\texttt{k}, n_2))
				\end{array}
				\right\}
			}
		\end{mathpar}
		\caption{Specification of \texttt{insertOp} for the give-up template}
		\label{fig:insertOp_giveup}
	\end{subfigure}
	\caption{Specification of \texttt{insertOp} for the lock-coupling and give-up templates}
	\label{fig:insertOp}
\end{figure*}

@@

\section{Proof Outlines for Templates (Sections \ref{traverse_proof_lock} and \ref{traverse_proof_giveup})}
\label{sec:apd_proof}
Below are proof outlines for the \texttt{traverse} function for the give-up template (Figure \ref{proof_giveup_traverse}) as well as the \texttt{insert} function for both templates. The complete proofs can be found in our Coq mechanization (see Section \ref{proof_mech}).
\begin{figure*}[h]
	$\color{specblue}
	\fforall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (p, n)  \ast  \nodeboxrep(n)  \big| \ \treerep\ (m)
	\end{array}
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
	$\color{specblue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
	\end{array}
	\ \Bigg| \ \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the give-up \texttt{traverse} function}
	\label{proof_giveup_traverse}
\end{figure*}

\begin{figure*}[h]
	\begin{subfigure}{\textwidth}
		\centering
		$\color{specblue}
		\fforall \  m. \left\langle
		\begin{array}{c}
			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\end{array}
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_insert.c}
		$\color{specblue}
		\left\langle
		\begin{array}{c}
			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\end{array}
		\right\rangle$
		\caption{Proof outline of the lock-coupling \texttt{insert} function}
		\label{proof_lock_insert}	
	\end{subfigure}\qquad

	\begin{subfigure}{\textwidth}
		\centering
		$\color{specblue}
		\fforall \  m. \left\langle
		\begin{array}{c}
			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\end{array}
		\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_insert.c}
		$\color{specblue}
		\left\langle
		\begin{array}{c}
			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\end{array}
		\right\rangle$
		\caption{Proof outline of the give-up \texttt{insert} function}
		\label{proof_giveup_insert}
	\end{subfigure}
	\caption{Proof outlines for the \texttt{insert} function}
	\label{proof_lock_giveup_insert}
\end{figure*}
\end{document}
\endinput
%%
%% End of file `sample-sigplan.tex'.

>>>>

%But first, we need to understand how to specify the correctness of \lstinline{traverse} independently of the data structure operations (insert, lookup, etc.) it will be used to implement. The key is the concept of \emph{logical atomicity}~\cite{tada}, a highly compositional approach to specifying concurrent data structure operations. \wm{Should this go in background instead?}
We specify the concurrent behavior of search structures using logical atomicity~\cite{tada, iris, iris-vst-arxiv}, a separation logic technique that concisely defines the behavior of concurrent operations. A logically atomic triple has the form ${\color{specblue} \fforall a.\,\left\langle \texttt{P}_l\ |\ \texttt{P}_p(a) \right\rangle } \ \texttt{c}\ {\color{specblue}\left\langle \texttt{Q}_l\ |\ \texttt{Q}_p(a)\right\rangle}$, where $\texttt{P}_l$ and $\texttt{Q}_l$ are \emph{local} preconditions and postconditions, akin to a standard Hoare triple, while $\texttt{P}_p$ and $\texttt{Q}_p$ are \emph{public} preconditions and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, this says that \lstinline{c} is an operation on an abstract object (i.e., data structure) $a$, and its effect is to \emph{atomically} transform $a$ from a state satisfying $\texttt{P}_p$ to a state satisfying $\texttt{Q}_p$, with no intermediate states visible to any other thread. More precisely, the triple asserts that if $\texttt{P}_l$ holds true before a call to \lstinline{c} and $\texttt{P}_p$ is true for some value of $a$ in a shared state, then $\texttt{P}_p$ will continue to be true for some (possibly different) value of $a$ until the \emph{linearization point} of $\texttt{c}$, at which point $\texttt{Q}_p$ will become true atomically for the same value $a$ (and $\texttt{Q}_l$ will be true after $\texttt{c}$ ends).

@@

\begin{figure*}[!ht]
	$\color{specblue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
	\right\rangle$
		\begin{minipage}{0.9\textwidth}
			\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
		\end{minipage}
	$\color{specblue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

\begin{figure*}[h]
	\begin{subfigure}{\textwidth}
		\centering
		\begin{mathpar}
			{\color{specblue}
				\left\{ 
				\begin{array}{c}
					\ \texttt{pn->n} \mapsto \texttt{NULL}  
				\end{array}
				\right\}
			} \vspace{-0.85em} 
			\\ 
			\texttt{insertOp(pn, k, v)} 
			\vspace{-0.85em} 
			\\
			{\color{specblue}
				\left\{\exists \  l\ r\ \mathit{lk}_1\ \mathit{lk}_2.
				\begin{array}{l}
					\ \texttt{pn->n->t} \mapsto (\texttt{k}, \texttt{v}, l, r) \ \ast \ \\ 
					\ l \mapsto (\texttt{NULL}, \mathit{lk}_1)  \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2)
				\end{array}
				\right\}
			}
		\end{mathpar}
		\caption{Specification of \texttt{insertOp} for the lock-coupling template}
		\label{fig:insertOp_lock}
	\end{subfigure}
	%insertOp for give-up of give-up
	\begin{subfigure}{\textwidth}
		\begin{mathpar}
			{\color{specblue}
				\left\{ 
				\begin{array}{c}
					\texttt{pn->n}\mapsto \texttt{NULL}  \ast  \texttt{pn->n->min}\mapsto n_1  \ast  \texttt{pn->n->max}\mapsto n_2
				\end{array}
				\right\}
			} \vspace{-0.85em} 
			\\
			\texttt{insertOp(pn, k, v)}
			\vspace{-0.85em} 
			\\
			{\color{specblue}
				\left\{\exists \ l\ r\ \mathit{lk}_1\ \mathit{lk}_2. 
				\begin{array}{c}
					\texttt{pn->n->t}\mapsto (\texttt{k}, \texttt{v}, l, r)  \ast  \texttt{pn->n->min}\mapsto n_1 \ast  \texttt{pn->n->max}\mapsto n_2  \ \ast 
					\\ l \mapsto (\texttt{NULL}, \mathit{lk}_1, (n_1, \texttt{k})) \ast  r \mapsto (\texttt{NULL}, \mathit{lk}_2, (\texttt{k}, n_2))
				\end{array}
				\right\}
			}
		\end{mathpar}
		\caption{Specification of \texttt{insertOp} for the give-up template}
		\label{fig:insertOp_giveup}
	\end{subfigure}
	\caption{Specification of \texttt{insertOp} for the lock-coupling and give-up templates}
	\label{fig:insertOp}
\end{figure*}

@@

\section{Proof Outlines for Templates (Sections \ref{traverse_proof_lock} and \ref{traverse_proof_giveup})}
\label{sec:apd_proof}
Below are proof outlines for the \texttt{traverse} function for the give-up template (Figure \ref{proof_giveup_traverse}) as well as the \texttt{insert} function for both templates. The complete proofs can be found in our Coq mechanization (see Section \ref{proof_mech}).
\begin{figure*}[h]
	$\color{specblue}
	\fforall \  m. \left\langle
	\begin{array}{c}
		\texttt{pn} \mapsto (p, n)  \ast  \nodeboxrep(n)  \big| \ \treerep\ (m)
	\end{array}
	\right\rangle$
	\begin{minipage}{0.9\textwidth}
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_traverse.c}
	\end{minipage}
	$\color{specblue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast \ 
		\\ 
		(\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range}))
	\end{array}
	\ \Bigg| \ \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the give-up \texttt{traverse} function}
	\label{proof_giveup_traverse}
\end{figure*}

\begin{figure*}[h]
	\begin{subfigure}{\textwidth}
		\centering
		$\color{specblue}
		\fforall \  m. \left\langle
		\begin{array}{c}
			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\end{array}
		\right\rangle$
		\begin{minipage}{0.9\textwidth}
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_insert.c}
		\end{minipage}
		$\color{specblue}
		\left\langle
		\begin{array}{c}
			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\end{array}
		\right\rangle$
		\caption{Proof outline of the lock-coupling \texttt{insert} function}
		\label{proof_lock_insert}	
	\end{subfigure}\qquad

	\begin{subfigure}{\textwidth}
		\centering
		$\color{specblue}
		\fforall \  m. \left\langle
		\begin{array}{c}
			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m)
		\end{array}
		\right\rangle$
		\begin{minipage}{0.9\textwidth}
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_giveup_insert.c}
		\end{minipage}
		$\color{specblue}
		\left\langle
		\begin{array}{c}
			\nodeboxrep(\texttt{r}) \ \big | \ \treerep\ (m[\texttt{k} \mapsto \texttt{v}])
		\end{array}
		\right\rangle$
		\caption{Proof outline of the give-up \texttt{insert} function}
		\label{proof_giveup_insert}
	\end{subfigure}
	\caption{Proof outlines for the \texttt{insert} function}
	\label{proof_lock_giveup_insert}
\end{figure*}
\end{document}
\endinput
%%
%% End of file `sample-sigplan.tex'.
